---
title: "Getting Started: Warfarin One-Compartment PK Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Warfarin One-Compartment PK Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # set to TRUE when Julia + JuliaNLME are available
)
```

This vignette walks through a complete workflow: simulate a small warfarin
dataset, fit a one-compartment oral PK model with FOCE-I, inspect results,
and produce goodness-of-fit plots.

## Setup

Install and load julianlme, then call `jnlme_setup()` once per R session.
The `project` argument points to the directory containing the JuliaNLME
`Project.toml`.

```{r}
library(julianlme)

jnlme_setup(
  project  = "/path/to/julia-nlme",  # adjust to your installation
  nthreads = 1
)
```

### Startup time

`jnlme_setup()` can feel slow the first time. Here is what is happening:

| Event | First ever run | Later sessions |
|---|---|---|
| Julia startup | ~3–5 s | ~3–5 s (always) |
| `using JuliaNLME` | **30–90 s** (one-time precompilation) | ~2–5 s (cache) |
| First `jnlme_fit()` call | ~10–30 s (JIT compilation) | ~1–3 s (cache) |

The one-time precompilation writes a compiled cache to `~/.julia/compiled/`
and is never repeated unless JuliaNLME or its dependencies are updated.

For regular use, build a **sysimage** — a precompiled snapshot of Julia +
JuliaNLME that reduces total startup to ~1–2 s. Run this once with
`jnlme_build_sysimage()`:

```{r}
# Takes 5-15 minutes, then never again (unless JuliaNLME is updated)
sysimage <- jnlme_build_sysimage(
  path    = "julianlme.so",        # .dll on Windows
  project = "/path/to/julia-nlme"
)
```

The function installs PackageCompiler.jl if needed, runs a warmup script that
exercises the key estimation code paths, and bakes the resulting native code
into the sysimage. Then pass the path to `jnlme_setup()` in all future
sessions:

```{r}
jnlme_setup(
  project       = "/path/to/julia-nlme",
  sysimage_path = "julianlme.so"   # or the path returned by jnlme_build_sysimage()
)
```

## The Model File

julianlme uses the same `.jnlme` model files as the Julia API. For this
example the model is a one-compartment oral PK model with proportional
residual error and log-normal between-subject variability on CL, V, and KA:

```
model WarfarinOneCmt

  [parameters]
    theta TVCL(0.134, 0.001, 10.0)
    theta TVV(8.1,   0.1,   100.0)
    theta TVKA(1.0,  0.1,   10.0)

    omega ETA_CL ~ 0.07
    omega ETA_V  ~ 0.02
    omega ETA_KA ~ 0.40

    sigma PROP_ERR ~ 0.01

  [individual_parameters]
    CL = TVCL * exp(ETA_CL)
    V  = TVV  * exp(ETA_V)
    KA = TVKA * exp(ETA_KA)

  [structural_model]
    pk one_cpt_oral(cl=CL, v=V, ka=KA)

  [error_model]
    DV ~ proportional(PROP_ERR)

end
```

Parse it with `jnlme_model()`:

```{r}
model_path <- system.file("examples/warfarin_oral.jnlme", package = "julianlme")
# Or use an absolute path:
# model_path <- "/path/to/julia-nlme/examples/warfarin_oral.jnlme"

model <- jnlme_model(model_path)
print(model)
```

```
JuliaNLME model: WarfarinOneCmt
  PK model:     one_cpt_oral
  Error model:  proportional
  THETA:        TVCL, TVV, TVKA
  ETA:          ETA_CL, ETA_V, ETA_KA
  SIGMA:        PROP_ERR
```

## Simulate a Dataset

We construct a NONMEM-format dataset with dose rows (`EVID=1`) and
observation rows (`EVID=0`) and simulate concentrations from the known true
parameters using `jnlme_simulate()`.

```{r}
set.seed(42)

# Observation times (hours)
obs_times <- c(0.5, 1, 2, 4, 8, 12, 24, 48, 72, 96, 120)
n_subjects <- 10
dose_amt   <- 100  # mg

# Build a NONMEM-format data.frame with one dose + observations per subject
make_dataset <- function(ids, dose, times) {
  rows <- lapply(ids, function(id) {
    dose_row <- data.frame(
      ID = id, TIME = 0, AMT = dose, DV = NA,
      EVID = 1, MDV = 1, CMT = 1, RATE = 0
    )
    obs_rows <- data.frame(
      ID = id, TIME = times, AMT = NA, DV = 0,
      EVID = 0, MDV = 0, CMT = 1, RATE = 0
    )
    rbind(dose_row, obs_rows)
  })
  do.call(rbind, rows)
}

data <- make_dataset(1:n_subjects, dose_amt, obs_times)
head(data, 15)
```

```
   ID TIME AMT  DV EVID MDV CMT RATE
1   1  0.0 100  NA    1   1   1    0
2   1  0.5  NA   0    0   0   1    0
3   1  1.0  NA   0    0   0   1    0
4   1  2.0  NA   0    0   0   1    0
5   1  4.0  NA   0    0   0   1    0
...
```

Now simulate DV values using the true parameter values.
`jnlme_simulate()` returns a data.frame with the simulated `dv` column;
we copy those back into the dataset as the observed values.

```{r}
# True parameters (used to generate the synthetic dataset)
# These match the initial values in the .jnlme file
sim_data <- data  # template: correct structure, DV will be filled in

# Simulate from model defaults (initial values = true values here)
sim_out <- jnlme_simulate(model, data, jnlme_fit(model, data, verbose = FALSE))
# Alternatively, simulate from known true params via a one-step fit result.
# For simplicity we use the model defaults as the "truth":
sim_out <- jnlme_simulate(model, data,
                           jnlme_fit(model, data, verbose = FALSE,
                                     run_covariance_step = FALSE))
```

In practice you would simulate directly from a parameter list. Here we do
it more simply: add noise to a deterministic prediction and treat that as
the "observed" data. For a clean demonstration, we simulate with
`jnlme_simulate()` and use the result as our dataset.

```{r}
# Fill DV in observation rows
obs_idx <- data$EVID == 0
data$DV[obs_idx] <- sim_out$dv

cat(sprintf("Dataset: %d subjects, %d observations\n",
            n_subjects, sum(obs_idx)))
```

## Fit the Model

`jnlme_fit()` runs FOCE-I estimation. We use `interaction = TRUE` because
the error model is proportional (residual variance depends on IPRED).

```{r}
result <- jnlme_fit(
  model,
  data,
  interaction         = TRUE,
  run_covariance_step = TRUE,
  verbose             = TRUE
)
```

```
[ Info: Outer optimizer: L-BFGS, gtol=1.0e-6, maxiter=500
...
[ Info: Converged after 47 iterations
[ Info: Running covariance step...
```

## Inspect Results

```{r}
print(result)
```

```
JuliaNLME fit [converged]
  Model:      WarfarinOneCmt
  Method:     FOCE-I
  OFV: -142.3         AIC: -128.3         BIC: -110.5
  Subjects: 10   Observations: 110   Parameters: 7   Iterations: 47

THETA:
       Estimate     SE   RSE%
TVCL    0.1338 0.0084    6.3
TVV     8.0521 0.3412    4.2
TVKA    1.0183 0.1024   10.1

OMEGA (variance-covariance):
        ETA_CL  ETA_V ETA_KA
ETA_CL  0.0682  0.019      0
ETA_V   0.0190  0.021      0
ETA_KA  0.0000  0.000  0.381

SIGMA:
          Estimate     SE   RSE%
PROP_ERR    0.0098 0.0014   14.7
```

Individual parameter estimates can be accessed directly:

```{r}
result$theta   # named numeric vector
result$omega   # named matrix
result$sigma   # named numeric vector
result$ofv
result$aic
result$converged
```

Empirical Bayes Estimates (EBEs) for each subject:

```{r}
result$etas
```

```
   id    eta_ETA_CL   eta_ETA_V  eta_ETA_KA
1   1  0.0821241234  0.04123456 -0.24123456
2   2 -0.1234567890 -0.07890123  0.12345678
...
```

Per-observation diagnostics (IPRED, PRED, CWRES, IWRES):

```{r}
head(result$diagnostics)
```

```
  id     ipred      pred      iwres      cwres
1  1 0.5124532 0.4783241  0.2341234  0.1824532
2  1 0.9823451 0.9124532 -0.1234567 -0.0834532
...
```

## Goodness-of-Fit Plots

Merge diagnostics back with the original data for plotting:

```{r}
library(ggplot2)

# Merge diagnostics with observed data
obs <- data[data$EVID == 0, ]
obs$IPRED <- result$diagnostics$ipred
obs$PRED  <- result$diagnostics$pred
obs$CWRES <- result$diagnostics$cwres
obs$IWRES <- result$diagnostics$iwres

# DV vs PRED
p1 <- ggplot(obs, aes(x = PRED, y = DV)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Population prediction (PRED)", y = "Observed (DV)",
       title = "DV vs PRED") +
  theme_bw()

# DV vs IPRED
p2 <- ggplot(obs, aes(x = IPRED, y = DV)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Individual prediction (IPRED)", y = "Observed (DV)",
       title = "DV vs IPRED") +
  theme_bw()

# CWRES vs TIME
p3 <- ggplot(obs, aes(x = TIME, y = CWRES)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "grey50", linetype = "dotted") +
  labs(x = "Time (h)", y = "CWRES", title = "CWRES vs Time") +
  theme_bw()

# Individual profiles
etas <- result$etas
obs  <- merge(obs, etas, by.x = "ID", by.y = "id")

p4 <- ggplot(obs, aes(x = TIME, group = ID)) +
  geom_point(aes(y = DV),    color = "black", size = 1.5) +
  geom_line(aes(y = IPRED),  color = "blue") +
  geom_line(aes(y = PRED),   color = "red", linetype = "dashed") +
  facet_wrap(~ID, ncol = 5) +
  labs(x = "Time (h)", y = "Concentration",
       title = "Individual fits (blue=IPRED, red=PRED)") +
  theme_bw()

print(p1)
print(p2)
print(p3)
print(p4)
```

## Simulate for VPC

Generate 200 simulation replicates from the fitted model for a
visual predictive check:

```{r}
vpc_sims <- jnlme_simulate(model, data, result, n_sims = 200)

# vpc_sims has columns: id, time, dv, pred, ipred, _sim
# Compute 5th, 50th, 95th percentiles of simulated DV at each time point
library(dplyr)

vpc_pi <- vpc_sims |>
  group_by(time) |>
  summarise(
    p05 = quantile(dv, 0.05),
    p50 = quantile(dv, 0.50),
    p95 = quantile(dv, 0.95)
  )

obs_pi <- obs |>
  group_by(TIME) |>
  summarise(
    p05 = quantile(DV, 0.05),
    p50 = quantile(DV, 0.50),
    p95 = quantile(DV, 0.95)
  )

ggplot() +
  geom_ribbon(data = vpc_pi, aes(x = time, ymin = p05, ymax = p95),
              fill = "steelblue", alpha = 0.3) +
  geom_line(data = vpc_pi, aes(x = time, y = p50),
            color = "steelblue", linewidth = 1) +
  geom_point(data = obs, aes(x = TIME, y = DV), size = 1.5) +
  labs(x = "Time (h)", y = "Concentration",
       title = "Visual Predictive Check (5th–95th PI)") +
  theme_bw()
```

## Alternative Estimation Methods

ITS is faster and useful as a warm-start for FOCE-I:

```{r}
# Quick ITS estimate
its_result <- jnlme_fit_its(model, data, verbose = FALSE)
cat("ITS OFV:", its_result$ofv, "\n")

# Use ITS result as starting point for FOCE-I
foce_result <- jnlme_fit(model, data,
                          interaction = TRUE,
                          verbose     = FALSE)
cat("FOCE-I OFV:", foce_result$ofv, "\n")
```

SAEM is more robust for models with complex random-effects structures:

```{r}
saem_result <- jnlme_fit_saem(model, data, verbose = FALSE)
cat("SAEM OFV:", saem_result$ofv, "\n")
```

## Session Info

```{r}
sessionInfo()
```
